Haskell in Computational Biology

### 2015-11-29 jfp.tex
Added MRFy journal paper to bib, correct typo

\pagetuningtrue % trying to get page breaks in the best spots

@@

\renewcommand\bibsection{\section*{\refname}}

@@

%
% It can't hurt to remind everyone involved exactly what an Experience
% Report is.
%
%
\begin{abstract}
%An Experience Report provides {evidence} that functional
%programming really works, or it describes obstacles that prevent
%functional programming from working.   
Haskell gives computational biologists
the flexibility and rapid prototyping of a scripting
language, and it compiles to native code.
From experience using Haskell to implement new algorithms for
computational biology, we~report where Haskell really worked for us
and what obstacles prevented it from working better.
Higher-order functions, lazy evaluation, and
monads worked well, but
profiling and debugging presented obstacles.
And~because allocation costs are difficult to predict and control,
compilation to native code did not guarantee great performance.
Finally, our experience with Haskell libraries varied greatly:
memoization combinators and parallel-evaluation
strategies helped us a lot,
but other, nameless libraries mostly got in our way.
Despite the obstacles and the uncertain quality of some libraries,
Haskell's ecosystem
made it easy for us to explore and run new algorithms.
\end{abstract}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell.
Although we encountered some obstacles that kept Haskell from working
as well as we would have liked, our overall experience was positive:
Haskell really worked for us, and we believe it will work equally well
for other computational biologists.
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
and the Haskell code is slower.
But the Haskell code was easy to write,     % coherent subject!
clearly implements the underlying mathematics
 (\secref{viterbi}),
 was easy to parallelize,
and performs well enough
(\secref{perf}).
  And our new tool solves a problem that could not be solved by
  the~C++ tool which preceded~it.
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one area:  understanding and
improving
performance (\secrefpair{improving-viterbi}{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although the first two authors are computational
biologists
with little functional-programming experience,
Haskell made it easy for us to explore new research ideas.
By~contrast,
our group's C++ code has 
made it hard to explore
% been a good vehicle for exploring 
new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some kept the promise, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
We couldn't tell in advance which 
would be which
 (\secref{penumbra}).
\end{itemize}
A~preliminary version of this
paper \citep{daniels:haskell-in-bio:2012} was presented as 
an \emph{Experience Report} in ICFP~2012.
This~paper adds a new technical section (\secref{improving-viterbi})
which describes our experience 
refactoring and improving our implementation of Viterbi's 
\citeyearpar{Viterbi:1967hq}
algorithm.
We~have also simplified the stochastic search as well as other elements of
the code.
Finally, we~have improved and added to explanations, figures, and diagrams.

@@

Computational biologists detect homologies by building 
algorithms which, given a {query sequence}, %% of amino acids,
compare it with known proteins.
When the known proteins have amino-acid sequences that
are not too different from the query sequence, homology can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research software, MRFy (pronounced
``Murphy''), can detect homologies 
in amino-acid sequences that are only distantly related.

A~paper for the computational biology community, based on MRFy, has been
submitted to the ACM Conference on Bioinformatics, Computational Biology, and 
Biomedical Informatics.
That paper explores the effectiveness of various search strategies,
and it also
compares MRFy with several other software tools.
MRFy is more accurate than the other software tested.
MRFy is available at \url{mrfy.cs.tufts.edu}.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.
%

@@

Once a protein alignment is constructed, it~is used to train a
\emph{hidden Markov model}. 
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence.
A~protein whose query sequence has a higher probability is more likely to %
be homologous to the proteins in the alignment.
We~write a query sequence as $x_1, \ldots, x_{\scriptscriptstyle N}$,
where each $x_i$~is 
an amino-acid \emph{residue}.\footnote
{When an amino acid combines with other amino acids to form a protein, it gives up
two hydrogen atoms and an oxygen atom.
What's left is called
a \emph{residue}.
In~this paper we use the terms ``residue'' and ``amino acid'' 
interchangeably.
}
The number of residues, $N$, can differ from the number of columns
in the alignment,~\alignwidth.

@@

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has residue~$x$ in
 column~$j$.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained residue~$x$ by
insertion at column~$j$.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that a homologous protein
has lost a~residue by deletion from column~$j$.
A~deletion state contains no tables of probabilities.
\hfuzz=1.5pt % fix blot
\end{itemize}
The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are \emph{emission probabilities}.
The names of the states are easily confused;
one~way to keep them straight is to imagine that the model of the
alignment describes a prototype sequence of length~$\alignwidth$,
and that each state describes an action used to build the query
sequence from the prototype: 
match a prototypical residue,
insert an unexpected residue,
or
delete a prototypical residue.

A~hidden Markov model also has distinguished ``begin'' and ``end''
states; begin and end states are like match states, but they
correspond to no residues and contain no emission probabilities.
%%
%% for consistency with the text above 'begin' and 'end' are neither
%% capitalized nor italicized.
%%
In~our representation, 
the begin and end states are known by their positions:
in a $\alignwidth$-column model, the begin state shares
column~0 with an insertion state, and the end state is the only state
in column~$\alignwidth$. 
Every other state is referred to by its column~$j$
and by one of these labels:\label{code:state-label}
\smallverbatiminput{statelabel}
We~group states that share a column~$j$; these states are referred
to collectively as ``node~$j$,'' as shown in \figref{plan7}.
This model
has begin and end states $B$~and~$E$;
an~insertion state in node~0; and
four
full nodes, each containing 
an insertion state~$I$, 
a match state~$M$, and a
deletion state~$D$.
(Node~0 may also be viewed as a full node, but its deletion state is
not reachable from the begin state, so it is shown with only two
states.)

@@

The model in \figref{plan7} is from a well-established family
that permits
transitions between match states and all other adjacent states,
but forbids
direct
transitions between insertion 
states and deletion states  \cite{Eddy:1998ut}. 
Models in this family are called ``Plan7'' models because
every full node has 7~transitions.
Each transition has its own probability; the probabilities for
transitions leaving states in node~$j$ are
$\txprobjj I I$,
$\txprobxj I M$,
$\txprobjj M I$,
$\txprobxj M M$,
$\txprobxj M D$,
$\txprobxj D M$,
and
$\txprobxj D D$.
(The transitions from match to insertion and from insertion to
insertion leave states in 
node~$j$ but also enter states in node~$j$.
The other transitions enter states in node~$j+1$.)
When column $j+1$ is a consensus column, transitions that enter
its match state 
%($\txprobxj I M$,
%$\txprobxj M M$,
%$\txprobxj D M$)
are more likely.
When column $j+1$ is a non-consensus column, transitions that enter
its insertion and deletion states are more likely.

@@

Given a hidden Markov model, 
an~established software package called HMMER (pronounced ``hammer'') 
can compute the probability
that a new protein shares structure 
with the proteins used to train the model.
The computation finds the most likely matching path through the hidden Markov model.
In~any matching path, each match state and insertion state corresponds to a
unique residue in the query sequence.
The probability of the path is the product of all of the corresponding
emission probabilities, together with the product of the transition
probabilities along the edges of the path.

Probabilities are small, and products of small numbers can become very
small indeed.
If~we were to multiply probabilities represented as floating-point
numbers, all the useful information would wind up in exponents,
and an exponent holds only
11~bits.
In~order to store useful information in the significand, which holds 53~bits,
we~follow standard practice: instead of multiplying probabilities,
we~add logarithms of probabilities \cite{Viterbi:1967hq}.
Also following standard practice, we~work with
\emph{negated}
logarithms of probabilities, which are  positive numbers.
The negated logarithm of a probability is called a \emph{score}:
\smallverbatiminput{score}
We~have given type \texttt{Score} a limited \texttt{Num} instance which permits
scores to be added and subtracted but not multiplied.

@@

The~most likely path through a model is the one with the
smallest score.
The~score of each path is computed by dynamic programming; the score associated with
a path to a state~$s$ ($M$, $I$, or~$D$) at node~$j$ and position~$i$ in
the query sequence is written $V_j^{\prime s}(i)$.
For~example, 
score~$V_j^{\prime M}(i)$ represents the probability of the most
likely path of the first $i$~residues in the query sequence,
terminating with placement of residue~$x_i$ in state~$M_j$.
Scores $V_j^{\prime I}(i)$~and~$V_j^{\prime D}(i)$ are similar.
Scores are specified by Viterbi's equations, which are shown
in \figref{viterbi-transformed}. 
The~equations (expressed using probabilities) are explained very clearly by
\citet[Chapter~5]{Durbin:1998wz}

To~use Haskell, we had to reimplement the standard
algorithm for solving Viterbi's equations.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.

\ifpagetuning\enlargethispage{1.2\baselineskip}\fi

Our code represents a query sequence as an immutable array of residues.
In~idiomatic Haskell, 
we might represent an individual residue~$x_i$
using a value of algebraic data type with constructors named for
amino acids:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | ...   -- not used
\end{smallverbatim}
But in our
legacy file format, each amino acid (residue) is a small integer
used only to index arrays.
%which we use only as an array index.
We therefore chose a representation that is easier to make consistent
with query sequences as represented on~disk:
\smallverbatiminput{aa}

@@

A~model is represented as a sequence of \emph{nodes}, 
each of which includes three states:\label{code:model3-node}
\smallverbatiminput{model3-node}
The representation of a state~$s$ includes the scores of transitions leaving
that state;
if~state~$s$ is a match or insertion state, its representation also
includes a table of emission scores.
For~example, the representation of a match state is \texttt{MState}:
\smallverbatiminput{model3-mstate}
%
The transition scores, which are shown as $a_{s\hat s}$
in \figref{plan7}, are read by  
function \texttt{aScore}, 
%which is given states $s$~and~$\hat s$ and a node index~$j$; its
whose
specification is
\mbox{$\mathtt{aScore}\;s\;\hat s\;j = \txprobxj s {\hat s}$}.\seclabel{aScore}
The
emission table is read by 
\texttt{eScore}, whose specification
is % parallel structure for line breaking below
\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$},
where $x_i$ is the residue located at position~$i$ in the query
sequence. 
%
%%  We place the transition probabilities 
%%  into a record in which each field is labeled
%%  \texttt{$s$\_$\mskip 1mu\hat s$},
%%  where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
%%  labels:\label{sec:tprobs}
%%  \smallverbatiminput{tprob-tprobs}
%%  \remark{Need to explain those last two fields. HMMer?}

@@

Function~\verb+vee'+ makes a recursive call not to itself but
to~\verb+vee''+. 
Function~\verb+vee''+ is the {memoized} version of~\verb+vee'+;
it~produces the same result as~\verb+vee'+,
but repeated calls return in constant time:
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
package
\path{Data.MemoCombinators}.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

@@

Beta strands require new equations and
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
\iffalse
depends on the amino acid~$x'$ in column~${\pairedwith i}$.
\else
depends on whatever amino acid~$x'$  is in column~${\pairedwith i}$.
\fi
If~$x'$ is in position~$i'$ in the query sequence, Viterbi's
equations are altered; for example,
$V_{j}^{\prime M}(i)$ depends not only on
$V_{j-1}^{\prime M}(i-1)$ but also on
$V_{\pairedwith j}^{\prime M}(i')$.
%%  \remark{I can't get an equation I
%%  believe in. ---NR I think it's fine, unless we want to introduce
%%  notation to distinguish paired amino acids from paired
%%  model nodes --ND}
The distance between $j$~and~$\pairedwith j$ can be as small as a few
columns or as large as a few hundreds of columns.
Because $V_j^{\prime M}(i)$~depends not only on nearby values but also on
$V_{\pairedwith j}^{\prime M}(i')$,
dynamic programming cannot compute the maximum likelihood quickly 
\cite{Menke:2010ti,Daniels:2012}.

@@

%
% N.B. very tight paragraph...
%
MRFy treats the beta 
strands in the model
 as ``beads'' which can slide along the query sequence.
A~positioning of the beta strands is called
 a \emph{placement}.
A~placement's likelihood is computed
based on frequencies of amino-acid pairs observed in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of the rest of
the query sequence, between and around beta strands,
is computed quickly and exactly 
using Viterbi's algorithm.
\iffalse
This likelihood is \emph{conditioned} on the placement.
\else
The exact result is a probability that is
\emph{conditioned} on the placement.
\fi

@@

\mrfy's search path follows only useful moves;
if a move is useless, 
\mrfy\ abandons it and moves again (in a new random direction) from the
previous point.
The~search is depicted graphically in \figref{search-moves}.
Solid lines represent useful moves; dashed lines represent
useless moves.
The~points are laid out vertically in the order in which they are
discovered;
as you move downward, time increases.
Horizontally, we~have organized the points into five columns; 
the bottom point in each column is useful, and
any points above it in the same column are useless.

@@

If~\mrfy\ could apply Hughes's technique, it~would search by composing
a~\emph{generator} 
that produces an infinite sequence of moves, 
a~\emph{filter} that selects the useful moves, and 
a~\emph{test} function that enumerates finitely many useful moves
and returns the final destination.
But~a~generator may produce an infinite sequence of useless moves.
For example, if \mrfy\ should stumble upon a global optimum, every move from
that point would be useless.
Given an infinite sequence of useless inputs, a~filter would not
produce any values, and the search would diverge.
This situation is depicted graphically in
\figref{filtered-search-moves}:
on the left-hand side, 
the first move is useless, the second move is useful, and all
subsequent moves are useless.
Filtering out useless moves would produce the sequence shown on the
right-hand side, which contains two points followed by a divergent
computation.

@@

Using these pieces, function \texttt{everyPt} produces an infinite
sequence containing a mix of useful and useless points.
Each useful point is tagged with cumulative cost and its score:
\smallfuzzverbatiminput{6pt}{everygen}
Both \texttt{nextPt} and \texttt{utility} are monadic, but 
we can still exploit laziness:
from its starting point, \texttt{everyPt} produces an infinite list of
randomly chosen successor points, then calls \texttt{costedUtility} to
tag each one with a cumulative cost and a utility.
Because we want to compose monadic computations in the same order we
compose ordinary functions, 
our code uses the monadic composition
operator~\texttt{=<<}, which is the bind operator with its arguments
swapped.

If \mrfy\ used Hughes's classic model of lazy search,
we~might expect to see an infinite sequence of points obtained by
applying \texttt{nextPt} to many points in succession, as~done by
the Prelude function \texttt{iterate}.
But~that's not how \mrfy\ works;
instead, 
the~infinite list \texttt{successors}
is computed by applying \texttt{nextPt} to
the \emph{single} initial point \texttt{startPt}, 
repeated infinitely many times.
Each repetition uses a
distinct source of random variation.

>>>>

\pagetuningfalse % trying to get page breaks in the best spots

@@

@@

%
% It can't hurt to remind everyone involved exactly what an Experience
% Report is.
%
%
\begin{abstract}
%An Experience Report provides {evidence} that functional
%programming really works, or it describes obstacles that prevent
%functional programming from working.   
Haskell gives computational biologists
the flexibility and rapid prototyping of a scripting
language, plus the performance of native code.
In~our experience, higher-order functions, lazy evaluation, and
monads really worked, but
profiling and debugging presented obstacles.
Also, Haskell libraries vary greatly:
memoization combinators and parallel-evaluation
strategies helped us a lot,
but other, nameless libraries mostly got in our way.
Despite the obstacles and the uncertain quality of some libraries,
Haskell's ecosystem
made it easy for us to develop new algorithms in computational
biology.
\end{abstract}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell.
Although we encountered some obstacles that kept Haskell from working
as well as we would have liked, our overall experience was positive:
Haskell really worked for us, and we believe it will work equally well
for other computational biologists.
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
and the Haskell code is slower.
But the Haskell code was easy to write,     % coherent subject!
clearly implements the underlying mathematics
 (\secref{viterbi}),
 was easy to parallelize,
and performs well enough
(\secref{perf}).
\ifpagetuning
  And our new tool solves a problem that could not be solved by
  the~C++ tool which preceded~it.
\else
  \ifpagetuning
  And overall,
  \else
  Moreover, measured as a whole, 
  \fi
  the software that incorporates our new
  code 
  performs significantly
  better than the~C++ software that preceded~it.
\fi
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one area:  understanding and
improving
performance (\secrefpair{improving-viterbi}{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although the first two authors are computational
biologists
with little functional-programming experience,
Haskell made it easy for us to explore new research ideas.
By~contrast,
our group's C++ code has 
made it hard to explore
% been a good vehicle for exploring 
new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some kept the promise, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
We couldn't tell in advance which 
\ifpagetuning
would be which
\else
experience we were likely to have
\fi
 (\secref{penumbra}).
\end{itemize}
A~preliminary version of this
paper \citep{daniels:haskell-in-bio:2012} was presented as 
an \emph{Experience Report} in ICFP~2012.
This~paper adds a new technical section (\secref{improving-viterbi})
which describes our experience 
refactoring and improving our implementation of Viterbi's 
\citeyearpar{Viterbi:1967hq}
algorithm.
We~have also simplified the stochastic search as well as other elements of
the code.
Finally, we~have improved and added to the figures and diagrams.

@@

Computational biologists detect homologies by building 
algorithms which, given a {query sequence}, %% of amino acids,
compare it with known proteins.
When the known proteins have amino-acid sequences that
are not too different from the query sequence, homology can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research software, MRFy (pronounced
``Murphy''), can detect homologies 
in amino-acid sequences that are only distantly related.
MRFy is available at \url{mrfy.cs.tufts.edu}.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.
%

@@

Once a protein alignment is constructed, it~is used to train a
\emph{hidden Markov model}. 
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence.
A~protein whose query sequence has a higher probability is more likely to %
be homologous to the proteins in the alignment.
We~write a query sequence as $x_1, \ldots, x_{\scriptscriptstyle N}$,
where each $x_i$~is 
an amino acid.
The number of amino acids, $N$, can differ from the number of columns
in the alignment,~\alignwidth.

@@

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that a homologous protein
has lost an amino acid by deletion from column~$j$.
A~deletion state contains no tables of probabilities.
\hfuzz=1.5pt % fix blot
\end{itemize}
The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are \emph{emission probabilities}.
The names of the states are easily confused;
one~way to keep them straight is to imagine that the model of the
alignment describes a prototype sequence of length~$\alignwidth$,
and that each state describes an action used to build the query
sequence from the prototype: 
match a prototypical amino~acid,
insert an unexpected amino~acid,
or
delete a prototypical amino~acid.

A~full hidden Markov model also has distinguished ``begin'' and ``end''
states; begin and end states are like match states, but they
correspond to no amino acids and contain no emission probabilities.
%%
%% for consistency with the text above 'begin' and 'end' are neither
%% capitalized nor italicized.
%%
In~our representation, 
the begin and end states are known by their positions:
in a $\alignwidth$-column model, the begin state is the only state in
column~0 and the end state is the only state in column~$\alignwidth$.
Every other state is referred to by its column~$j$
and by one of these labels:\label{code:state-label}
\smallverbatiminput{statelabel}
We~group states that share a column~$j$; these states are referred
to collectively as ``node~$j$,'' as shown in \figref{plan7}.
This model
has begin and end states $B$~and~$E$,
as well as four
nodes, each containing 
an insertion state~$I$, 
a match state~$M$, and a
deletion state~$D$.
\remark{We need to figure out if the begin state is really node 0}

@@

The model in \figref{plan7} is from a well-established family
that permits
transitions between match states and all other adjacent states,
but forbids
direct
transitions between insertion 
states and deletion states  \cite{Eddy:1998ut}. 
The family is called the ``Plan7'' family because
every model in the family has 7~transitions per node.
Each transition has its own probability; the probabilities for
transitions leaving states in node~$j$ are
$\txprobjj I I$,
$\txprobxj I M$,
$\txprobjj M I$,
$\txprobxj M M$,
$\txprobxj M D$,
$\txprobxj D M$,
and
$\txprobxj D D$.
(The transitions from match to insertion and from insertion to
insertion leave states in 
node~$j$ but also enter states in node~$j$.
The other transitions enter states in node~$j+1$.)
When column $j+1$ is a consensus column, the transitions that enter
its match state 
%($\txprobxj I M$,
%$\txprobxj M M$,
%$\txprobxj D M$)
are more likely.
When column $j+1$ is a non-consensus column, the transitions that enter
its insertion and deletion states are more likely.

@@

Given a hidden Markov model, 
an~established software package called HMMER (pronounced ``hammer'') 
can compute the probability
that a new protein shares structure 
with the proteins used to train the model.
The computation finds the most likely matching path through the hidden Markov model.
In~any matching path, each match state and insertion state corresponds to a
unique amino acid in the query sequence.
The probability of the path is the product of all of the corresponding
emission probabilities, together with the product of the transition
probabilities along the edges of the path.

Probabilities are small, and multiplying small numbers makes smaller
numbers.
Using floating-point arithmetic, multiplying probabilities amounts to
storing data in the exponent of a floating-point number, which holds
only 11~bits.
In~order to store data in the significand, which holds 53~bits,
we~instead sum the logarithms of probabilities \cite{Viterbi:1967hq}.
Moreover, 
 our models use a
legacy file format which
\emph{negates}
the logarithm of each probability, making it a positive number.
%%To~make these numbers more pleasant to read, we negate them,
%%which eliminates hordes of minus signs from our file formats.
The negated logarithm of a probability is called a \emph{score}.
\smallverbatiminput{score}
Type \texttt{Score} has a limited \texttt{Num} instance which permits
scores to be added and subtracted but not multiplied.

@@

The~most likely path through a model is the one with the
smallest score.
The~score of each path is computed by dynamic programming; the score associated with
a path to a state~$s$ ($M$, $I$, or~$D$) at node~$j$ and position~$i$ in
the query sequence is written $V_j^{\prime s}(i)$.
For~example, 
score~$V_j^{\prime M}(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
Scores $V_j^{\prime I}(i)$~and~$V_j^{\prime D}(i)$ are similar.
Scores are specified by Viterbi's equations, which are shown
in \figref{viterbi-transformed}. 
The~equations (expressed using probabilities) are explained very clearly by
\citet[Chapter~5]{Durbin:1998wz}

To~be able to use Haskell, we had to reimplement the standard
algorithm for solving Viterbi's equations.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.

Our code represents a query sequence as an immutable array of amino
acids.
In~idiomatic Haskell, 
we might represent an individual amino acid~$x_i$
using a value of algebraic data type:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | ...   -- not used
\end{smallverbatim}
But in our
legacy file format, each amino acid is a small integer
used only to index arrays.
%which we use only as an array index.
We therefore chose 
\smallverbatiminput{aa}

@@

A~model is represented as a sequence of \emph{nodes}, 
each of which includes three states:\label{code:model3-node}
\smallverbatiminput{model3-node}
The representation of a state~$s$ includes the scores of transitions leaving
that state;
if~state~$s$ is a match or insertion state, its representation also
includes a table of emission scores.
For~example, the representation of a match state is \texttt{MState}:
\smallverbatiminput{model3-mstate}
%
The transition scores, which are shown as $a_{s\hat s}$
in \figref{plan7}, are read by  
function \texttt{aScore}, 
%which is given states $s$~and~$\hat s$ and a node index~$j$; its
whose
specification is
\mbox{$\mathtt{aScore}\;s\;\hat s\;j = \txprobxj s {\hat s}$}.
The
emission table is read by 
\texttt{eScore}, whose specification
is % parallel structure for line breaking below
\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$},
where $x_i$ is the amino acid located at position~$i$ in the query
sequence. 
%
%%  We place the transition probabilities 
%%  into a record in which each field is labeled
%%  \texttt{$s$\_$\mskip 1mu\hat s$},
%%  where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
%%  labels:\label{sec:tprobs}
%%  \smallverbatiminput{tprob-tprobs}
%%  \remark{Need to explain those last two fields. HMMer?}

@@

Function~\verb+vee'+ makes a recursive call not to itself but
to~\verb+vee''+. 
Function~\verb+vee''+ is the {memoized} version of~\verb+vee'+;
it~produces the same result as~\verb+vee'+,
but faster: 
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
package
\path{Data.MemoCombinators}.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

@@

Beta strands require new equations and
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
\ifpagetuning
depends on the amino acid~$x'$ in column~${\pairedwith i}$.
\else
depends on whatever amino acid~$x'$  is in column~${\pairedwith i}$.
\fi
If~$x'$ is in position~$i'$ in the query sequence, Viterbi's
equations are altered; for example,
$V_{j}^{\prime M}(i)$ depends not only on
$V_{j-1}^{\prime M}(i-1)$ but also on
$V_{\pairedwith j}^{\prime M}(i')$.
%%  \remark{I can't get an equation I
%%  believe in. ---NR I think it's fine, unless we want to introduce
%%  notation to distinguish paired amino acids from paired
%%  model nodes --ND}
The distance between $j$~and~$\pairedwith j$ can be as small as a few
columns or as large as a few hundreds of columns.
Because $V_j^{\prime M}(i)$~depends not only on nearby values but also on
$V_{\pairedwith j}^{\prime M}(i')$,
dynamic programming cannot compute the maximum likelihood quickly 
\cite{Menke:2010ti,Daniels:2012}.

@@

%
% N.B. very tight paragraph...
%
MRFy treats the beta 
strands in the model
 as ``beads'' which can slide along the query sequence.
A~positioning of the beta strands is called
 a \emph{placement}.
A~placement's likelihood is computed
based on frequencies of amino-acid pairs observed in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of the rest of
the query sequence, between and around beta strands,
is computed quickly and exactly 
using Viterbi's algorithm.
\ifpagetuning
This likelihood is \emph{conditioned} on the placement.
\else
The exact result is a probability that is
\emph{conditioned} on the placement.
\fi

@@

\mrfy's search path follows only useful moves;
if a move is useless, 
\mrfy\ abandons it and moves again (in a new random direction) from the
previous point.
The~search is depicted graphically in \figref{search-moves}.
The~solid lines represent useful moves; the dotted lines represent
useless moves.
The~points are laid out vertically in the order in which they are
discovered;
if~you like, time increases as you move downward.
Horizontally, we~have organized the points into five columns; 
the bottom point in each column is the result of a useful move, and we
call it a useful point.
Any points above it in the same column are useless.

@@

If~\mrfy\ could apply Hughes's technique, it~would search by composing
a~\emph{generator} 
that produces an infinite sequence of moves, 
a~\emph{filter} that selects the useful moves, and 
a~\emph{test} function that enumerates finitely many useful moves
and returns the final destination.
But~a~generator may produce an infinite sequence of useless moves.
(For example, if \mrfy\ should stumble upon a global optimum, every move from
that point would be useless.)
Given an infinite sequence of useless inputs, a~filter would not
produce any values, and the search would diverge.
This situation is depicted graphically in
\figref{filtered-search-moves}:
on the left-hand side, 
the first move is useless; the second move is useful; and all
subsequent moves are useless.
Filtering out useless moves would produce the sequence shown on the
right-hand side, which contains two points followed by a divergent
computation.

@@

Using these pieces, function \texttt{everyPt} produces an infinite
sequence containing a mix of useful and useless points.
Each useful point is tagged with cumulated cost and its score:
\smallfuzzverbatiminput{6pt}{everygen}
Both \texttt{nextPt} and \texttt{utility} are monadic, but 
we can still exploit laziness:
from its starting point, \texttt{everyPt} produces an infinite list of
randomly chosen successor points, then calls \texttt{costedUtility} to
tag each one with a cumulative cost and a utility.
Because we want to compose monadic computations in the same order we
compose ordinary functions, 
our code uses the monadic composition
operator~\texttt{=<<}, which is the bind operator with its arguments
swapped.

If \mrfy\ used Hughes's classic model of lazy search,
we~could possibly expect to see an infinite sequence of points obtained by
applying \texttt{nextPt} to many points in succession, as~done by
the Prelude function \texttt{iterate}.
But~that's not how \mrfy\ works;
instead, 
the~infinite list \texttt{successors}
is computed by applying \texttt{nextPt} to
the \emph{single} initial point \texttt{startPt}, \mbox{using} infinitely
many distinct sources of random variation.
To~understand the structure of
 \mrfy's more complex search, it~may help to
relate the code to \figref{search-moves}.
Please assume that \texttt{startPt} is the point in the upper left
corner of \figref{search-moves}.
Then the first part of the \texttt{successors} list appears in the
next column of \figref{search-moves}:
the~first four elements are \texttt{useless};
the next element, which is the first useful point, is~\texttt{pt};
and~the remaining elements, which are never evaluated,
match the
underscore.
The first useful point is located by \mono{span~isUseless}.
After the search reaches~\texttt{pt},
we~start the search anew with a recursive 
call to \mono{everypt~sg}, 
passing~\texttt{pt}.
%
The most informative part of \texttt{everyPt} is last expression of
the \texttt{do} block,
which shows that the result begins with a useful point,
is followed by a (possibly infinite, possibly empty) list of useless
points, and then continues with the results of a recursive call
to \texttt{everyPt}.
